---
title: "Reproducible Research Course Project 1"
author: "Vipul Jha"
date: "13 August 2020"
output: html_document
---

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

* Dataset: [Activity monitoring data] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) 

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format 
interval: Identifier for the 5-minute interval in which measurement was taken 
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset. 

## Loading and preprocessing the data

```{r}
setwd("C:/Users/Vipul/Desktop/repdata_data_activity")
activityData <- read.csv(file="activity.csv", header=TRUE)
```

## What is mean total number of steps taken per day?

```{r}
# Calculate the total steps taken per day

totalSteps <- aggregate(steps ~ date, activityData, FUN=sum)

# A histogram of the total number of steps taken per day

hist(totalSteps$steps,
     main = "Total Steps per Day",
     xlab = "Number of Steps")
```

```{r}
# Calculate the mean and median of total steps taken per day

meanSteps <- mean(totalSteps$steps, na.rm = TRUE)
medSteps <- median(totalSteps$steps, na.rm = TRUE)
```

## What is the average daily activity pattern?
```{r}
# A time-series plot of the 5-minute interval and NO. of steps taken averaged across all days

library(ggplot2)
meanStepsByInt <- aggregate(steps ~ interval, activityData, mean)
ggplot(data = meanStepsByInt, aes(x = interval, y = steps)) +
  geom_line() +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Which 5-minute interval across all days contain the maximum number of steps?

maxInt <- meanStepsByInt[which.max(meanStepsByInt$steps),]
```

## Inputing Missing Values

```{r}
# Calculate the total number of missing values in the dataset

missingVals <- is.na(activityData$steps)
```

There are 17568 missing values. Replacing these missing values with the 5-day average of that respective interval.

```{r}
# Create a new dataset equal to the original dataset but with the missing data filled in

imp_activityData <- transform(activityData,
                   steps = ifelse(is.na(activityData$steps),
                  meanStepsByInt$steps[match(activityData$interval, 
                    meanStepsByInt$interval)],
                     activityData$steps))

# Make a histogram of the total number of steps taken each day and report the mean and median

impStepsByInt <- aggregate(steps ~ date, imp_activityData, FUN=sum)
hist(impStepsByInt$steps,
     main = "Imputed Number of Steps Per Day",
     xlab = "Number of Steps")
```

```{r}
impMeanSteps <- mean(impStepsByInt$steps, na.rm = TRUE)
impMedSteps <- median(impStepsByInt$steps, na.rm = TRUE)
diffMean = impMeanSteps - meanSteps
diffMed = impMedSteps - medSteps
diffTotal = sum(impStepsByInt$steps) - sum(totalSteps$steps)
```

Difference in the mean of the two datasets - 0
Difference in the median of the two datasets - -1.076381110^4
Difference in the total steps of the two datasets - 8.612950910^4

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Create a new factor variable in the dataset with two levels - "weekend" and "weekday"

DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
      return ("weekeday")
  else if (day %in% c('Saturday', 'Sunday'))
      return ("weekend")
  else
      stop ("Invalid Date Format.")
}

imp_activityData$date <- as.Date(imp_activityData$date)
imp_activityData$day <- sapply(imp_activityData$date, FUN = DayType)

# A panel plot containing a time-series plot of the 5-minute interval and the average number of steps taken across all weekdays or weekends

meanStepsByDay <- aggregate(steps ~ interval + day, imp_activityData, mean)
ggplot(data = meanStepsByDay, aes(x = interval, y = steps)) + 
  geom_line() +
  facet_grid(day ~ .) +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```













